---
title: "sam_yizhou_modeling"
format: html
---

```{r}
#| label: "setup"
#| message: FALSE

here::i_am("gc2/sam_yizhou_modeling.qmd")
library(conflicted)
library(here)
library(tidyverse)
library(scico)
library(sf)
library(rnaturalearth)
library(geoR)
conflicts_prefer(dplyr::filter, dplyr::lag)

data <- read.csv(here("data/processed/annual_bat_counts.csv")) 
data <- data |> 
    rename(ALL_count = ALL) |>
    # Drops the mapping polygon information
    select(
        # Station and spatial information
        AcousticSite, Lat, Long, dist_to_water, Year, 
        # Counts, number of nights station on, and species tracked
        ends_with("count"), starts_with("nights"), ends_with("meta")
    ) |>
    # Create new "nights" column, just corresponding to the value of nights_X, 
    # where X is the value of the "Year" column
    rowwise() %>%
    mutate(nights = get(paste0("nights_", Year))) %>%
    ungroup() |> 
    select(-starts_with("nights_"))

theme_set(theme_bw())
```

Now looking at the full dataset (not dropping stations), the ANTPAL, CORRAF, LASINT, TADBRA species are tracked, so we will no longer drop them for our subsequent analysis.

```{r}
data_long <- data |>
    group_by(AcousticSite, Lat, Long, dist_to_water, Year) |>
    pivot_longer(
        cols = ends_with(c("_count", "_meta")), 
        names_to = c("Species", ".value"),
        names_sep = "_"
    ) |>
    mutate(meta = if_else(Species == "ALL", "yes", meta)) |>
    filter(meta == "yes") |>
    select(-meta)
```

# Visualize the spatial distribution of mean sightings per night by species (Great Lakes region only)
```{r}
sightings_df <- data_long |>
    mutate(SightingsPerNight = count / nights)
```

It is difficult to discern any patterns in the spatial variation of the mean number of sightings per night by species.
```{r}
mainmap <- ne_states(country = c("united states of america", "canada"), returnclass = "sf")
michiganplus <- mainmap %>% dplyr::filter(name %in% c("Illinois", "Michigan", "Wisconsin", "Ontario", "Ohio", "Indiana", "Pennsylvania", "New York", "West Virginia"))

sightings_df |>
    ggplot() +
    geom_point(size = .8, aes(x = Long, y = Lat, fill = SightingsPerNight)) +
    geom_sf(data = michiganplus, fill = "white", color = "black") +
    coord_sf(ylim = c(41, 49), xlim = c(-77, -91)) +
    facet_wrap(~ Species, nrow = 2) +
    theme_minimal() +
    scale_fill_scico(palette="batlowK") +
    ggtitle("Mean Sightings Per Night at Each Station by Species (Great Lakes Region)")
```

A histogram of total sightings by station suggests outliers for >= 100 sightings per night.
```{r}
hist(sightings_df$SightingsPerNight)
```

The lack of a clear spatial pattern continues even when we filter out seeming outliers in the number of mean sightings per night.
```{r}
sightings_df |>
    filter(SightingsPerNight <= 100) |>
    ggplot() +
    geom_point(size = .8, aes(x = Long, y = Lat, fill = SightingsPerNight)) +
    geom_sf(data = michiganplus, fill = "white", color = "black") +
    coord_sf(ylim = c(41, 49), xlim = c(-77, -91)) +
    facet_wrap(~ Species, nrow = 2) +
    theme_minimal() +
    scale_fill_scico(palette="batlowK") +
    ggtitle("Mean Sightings Per Night at Each Station by Species (Great Lakes Region)", subtitle = "Sightings/night >100 removed")
```

# Visualize the empirical semivariogram of total sightings by species

```{r}
#| message: FALSE
#| 
sv_data <- data.frame(
    species = character(), dists = numeric(), variogram = numeric(), npairs = numeric(), sd = numeric()
)
for(sp in unique(data_long$Species)){
    sp_data <- sightings_df |> filter(Species == sp)
    # Plot the semivariogram for each species
    sv <- sp_data %>% with(geoR::variog(
      data = SightingsPerNight, coords = cbind(Long, Lat), 
      uvec = 30, messages=FALSE
    ))
    sv_df <- data.frame(species = sp, dists = sv$u, variogram = sv$v, npairs = sv$n, sd = sv$sd)
    sv_data <- rbind(sv_data, sv_df)
}
ggplot(sv_data, aes(x=dists, y=variogram)) + 
    geom_point(size=2, shape=8) +
    facet_wrap(~ species, nrow = 2, scales = "free") +
    theme_minimal() + 
    xlab("Distances") +
    ylab("Variogram")
```