---
title: "sam_yizhou_eda"
format: html
---

```{r}
#| label: "setup"
#| message: FALSE

here::i_am("gc1/sam_yizhou_eda.qmd")
library(conflicted)
library(here)
library(tidyverse)
library(scico)
library(sf)
library(rnaturalearth)
library(geoR)
conflicts_prefer(dplyr::filter, dplyr::lag)

load(here("data/processed/data_metadata.RData"))

theme_set(theme_bw())
```

Note that the ANTPAL, CORRAF, LASINT, TADBRA species are not tracked in any stations, so we will drop them for our subsequent analysis.

```{r}
metadata$ANTPAL |> table()
metadata$CORRAF |> table()
metadata$LASINT |> table()
metadata$TADBRA |> table()

miss_spec <- c("ANTPAL", "CORRAF", "LASINT", "TADBRA")

vis_data <- data |> 
    select(-one_of(miss_spec)) 
species <- setdiff(names(vis_data), c("AcousticSite", "NightOf", "ALL", miss_spec))

vis_metadata <- metadata |> 
    select(-one_of(miss_spec)) |>
    group_by(AcousticSite, Lat, Long, across(all_of(species))) |>
    summarise(detectorNights = sum(across(all_of(as.character(2010:2018))))) |>
    ungroup()
vis_metadata2 <- vis_metadata |> 
    select(AcousticSite, Lat, Long, detectorNights)
```

# Visualize the spatial distribution of mean sightings per night by species
```{r}
sightings_df <- vis_data |>
    select(-NightOf) |>
    group_by(AcousticSite) |>
    summarise(across(everything(), sum)) |>
    inner_join(vis_metadata2, "AcousticSite") |>
    pivot_longer(
        cols = all_of(species), names_to = "Species", values_to = "TotalSightings"
    ) |>
    mutate(SightingsPerNight = TotalSightings / detectorNights)
```

It is difficult to discern any patterns in the spatial variation of the total number of sightings by species.
```{r}
mainmap <- ne_states(country = c("united states of america", "canada"), returnclass = "sf")
michiganplus <- mainmap %>% dplyr::filter(name %in% c("Illinois", "Michigan", "Wisconsin", "Ontario", "Ohio", "Indiana", "Pennsylvania", "New York", "West Virginia"))

sightings_df |>
    ggplot() +
    geom_point(size = .8, aes(x = Long, y = Lat, fill = SightingsPerNight)) +
    geom_sf(data = michiganplus, fill = "white", color = "black") +
    coord_sf(ylim = c(41, 49), xlim = c(-77, -91)) +
    facet_wrap(~ Species, nrow = 2) +
    theme_minimal() +
    scale_fill_scico(palette="batlowK") +
    ggtitle("Mean Sightings Per Night at Each Station by Species")
```

A histogram of total sightings by station suggests outliers for >= 100 sightings per night.
```{r}
hist(sightings_df$SightingsPerNight)
```

The lack of a clear spatial pattern continues even when we filter out seeming outliers in the number of total sightings.
```{r}
sightings_df |>
    filter(SightingsPerNight <= 100) |>
    ggplot() +
    geom_point(size = .8, aes(x = Long, y = Lat, fill = SightingsPerNight)) +
    geom_sf(data = michiganplus, fill = "white", color = "black") +
    coord_sf(ylim = c(41, 49), xlim = c(-77, -91)) +
    facet_wrap(~ Species, nrow = 2) +
    theme_minimal() +
    scale_fill_scico(palette="batlowK") +
    ggtitle("Mean Sightings Per Night at Each Station by Species", subtitle = "(Sightings/night >100 removed)")
```

# Visualize the empirical semivariogram of total sightings by species

To create our semivariogram for each species, we have to merge the data and metadata to get the Latitude and Longitude information for the stations that track the species of interest.
```{r}
#| message: FALSE
#| 
sv_data <- data.frame(
    species = character(), dists = numeric(), variogram = numeric(), npairs = numeric(), sd = numeric()
)
for(sp in species){
    # Metadata for the stations relevant to the species of interest
    vis_metadata_sub <- vis_metadata |> 
        filter(.data[[sp]] == "yes") |>
        select(AcousticSite, Long, Lat, detectorNights) 
    # Get total sightings across years for each station
    sp_data <- inner_join(vis_data, vis_metadata_sub, "AcousticSite") |>
        group_by(AcousticSite, Long, Lat, detectorNights) |>
        summarise(TotalSightings = sum(.data[[sp]])) |>
        mutate(SightingsPerNight = TotalSightings / detectorNights)
    # Plot the semivariogram for each species
    sv <- sp_data %>% with(geoR::variog(data = SightingsPerNight, coords = cbind(Long, Lat), messages=FALSE))
    sv_df <- data.frame(species = sp, dists = sv$u, variogram = sv$v, npairs = sv$n, sd = sv$sd)
    sv_data <- rbind(sv_data, sv_df)
}
ggplot(sv_data, aes(x=dists, y=variogram)) + 
    geom_point(size=2, shape=8) +
    facet_wrap(~ species, nrow = 2, scales = "free") +
    theme_minimal() + 
    xlab("Distances") +
    ylab("Variogram")
```